---
title: "Appendix 3: World Maps"
author: 'Chris Bentz'
date: "July 20, 2022"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Session Info
Give the session info (reduced).
```{r, echo = F}
# R version
sessionInfo()$R.version$version.string
# platform
sessionInfo()$R.version$platform 
```

# Load Libraries
If the libraries are not installed yet, you need to install them using, for example, the command: install.packages("ggplot2").
```{r, message = FALSE}
library(readr)
library(ggmap)
library(maps)
library(gridExtra)
library(ggrepel)
```

Give the package versions.
```{r, echo = F}
# version of packages loaded
installed.packages()[names(sessionInfo()$otherPkgs), "Version"]
```

# Load the Data 
Load Glottolog (Version 4.1) language information combined with information on the language sample of the IWMLC.
```{r, message = F}
languages <- as.data.frame(read_csv("https://raw.githubusercontent.com/IWMLC/language-complexity-metrics/master/languages_metaInfo.csv"))
```

# Simple Stats
```{r}
length(unique(languages$isocodes)) # number of languages according to iso
length(unique(languages$glottocode)) # number of languages according to glottolog
length(unique(languages$family_id)) # number of language families according to glottolog
unique(languages$macroarea) # number of macroareas according to glottolog
```

# Pre-Processing
```{r}
# remove Norwegian Nynorsk (nno) since this has NAs in glottolog
languages <- languages[languages$isocodes != "nno", ]
# add 360 to longitudes for languages with longitudes < -25 
# (this is necessary to create a Pacific centered map)
languages$longitude[languages$longitude < -25] <- 
  languages$longitude[languages$longitude < -25] + 360
```

# World Map
World maps with macroarea information from Glottolog.
```{r, fig.width = 9, fig.height = 9, warning = FALSE}
# create world map
world <- map_data("world", wrap = c(-25, 335)) 
area.map <- ggplot() + 
  geom_polygon(data = world, aes(x = long, y = lat, group = group), 
               fill = "white", colour = "darkgrey") +
  geom_point(data = languages, aes(x = longitude, y = latitude, 
                                 fill = macroarea), 
             alpha = 0.9, size = 2.5, pch = 21) +
  geom_text_repel(data = languages, aes(x = longitude, y = latitude, 
                                        label = name), size = 2.5,
                  box.padding = unit(0.1, 'lines'), force = 0.2) +
  scale_y_continuous(limits = c(-65, 80)) +
  labs(x = "longitude", y = "latitude", fill = "Macroarea") +
  theme_bw() +
  facet_wrap (~ corpus, nrow = 2) +
  theme(axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        title = element_text(size = 12),
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.position = "bottom")
area.map
```

Save to file.
```{r, warning = FALSE}
ggsave("Figures/WorldMap/worldMaps.pdf", area.map, 
       dpi = 300, scale = 1, width = 9, height = 9, device = cairo_pdf)
```
